---
title: "Introduction to R Markdown"
subtitle: Exercise 1
output: html_document
author: your name
params: 
  color: "blue"
---

This notebook is meant as a small playground to get more comfortable
with using R Markdown. For a quick overview of the most important functionalties 
check out the R Markdown [cheat sheet](https://rmarkdown.rstudio.com/lesson-15.html).
The cheat sheet is also available under (*Help -> Cheat Sheets*) in RStudio. For more in depth information the official R Markdown site might be a good starting point [https://rmarkdown.rstudio.com](https://rmarkdown.rstudio.com).

## R Markdown fundamentals

Produce an output document from the R Markdown file 
by clicking on the *Knit* button. Run the code chunk below.

```{r cars}
str(cars)
```

Insert and run a new code chunk to calculate the mean speed and mean stopping distance of the cars.

```{r}
colMeans(cars)
```


## Markdown Formatting

Create an unordered list containing three git commands. Make the first item bold, the second italic and display the last item as inline code.

* **git status**
* *git add*
* `git commit`

Insert a link to http://example.edu with the caption text *example*.

[example](http://example.edu)

Write down the Poisson distribution using inline math.

$P_{\lambda}(k) = \frac{\lambda^k}{k!} e^{-\lambda}$

## Chunk Options

Insert a new code chunk containing `3 * 2` and name it *Simple Calculation*.

```{r simple calculation}
3 * 2
```

Create a new chunk called setup and use it to load the `ggplot2` package. Ensure that the
code chunk and its output is not included in the final document.

```{r setup, include = FALSE}
library(ggplot2)
```

Plot a scatter plot of car `dist` against `speed`. Prevent the code from being displayed in the output document.

```{r, echo = FALSE}
ggplot(data = cars) + geom_point(mapping = aes(speed, dist))
```

## Inline Code

You can add inline code into an R Markdown file. Fill in the gaps of the sentence below
with the appropriate inline code.

The `cars` data frame contains `r nrow(cars)` observations. The highest speed recorded in
the data frame is `r max(cars$speed)` mph.


## YAML Header

Set the default output as PDF or HTML document in the YAML header. Note that producing a PDF will require LaTeX to be installed on your computer.

You can use the YAML header to define parameters for your report. This is especially useful, if you might need to produce a similar report multiple times. Declare a parameter named `color` in the YAML header, and assign a sensible value to it. Add a new code chunk containing the same scatter plot from above, but
make the color of the data points dependent on the parameter value set in the YAML header.



```{r, echo = FALSE}
ggplot(data = cars) + geom_point(mapping = aes(speed, dist), color = params$color)
```


